Bitmap Indexing for Fast Analysis of Large Astrophysical Datasets

### 2016-04-01 ms/ms.tex
Complete read through edits

%---------------------------------------------------------------------------------
% ABSTRACT
\begin{abstract}
Current scientific datasets can reach unwieldy sizes that impede analysis and scientific discovery. We present a method for using Morton order bitmap indexes to map the parts of the full domain touched by files within a dataset. This takes advantage of domain partitioning across multiple files and limits the volume of data that must be loaded into memory for a single query. This method is tested on mock datasets containing $1024^3$ three-dimensional points split across 512 files according to different partitioning schemes. Using Morton order bitmap indexes for localized partitioning schemes like the Hilbert space-filling curve, the number of files that must be loaded to access data for a sub-section of the domain will scale with the size of the sub-section and can be identified in under 1\,s even when the entire domain is included. This will dramatically decrease the time and memory required for loading data during analysis.
\end{abstract}

@@

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% INTRO
\section{Introduction}\label{S:intro}
Continuing advances in both hardware and software facilitate astrophysical datasets of growing complexity and size. The datasets produced by numerical simulations can currently reach sizes of a $\sim$100\TB{ }split across hundreds of files \citep[e.g.][]{Croft2015}. This problem is not limited to theoretical work. During operations the Large Synoptic Survey Telescope (LSST) will produce 15\TB{ }of data each night \citep{Juric2015a}. In order to analyze such large datasets, we need innovative techniques for quickly indexing and selecting data without loading the entire dataset into memory. We present a technique for using Morton bitmap indexes to map files and accelerate data analysis. \S\ref{S:theory} provides background information on Morton order indexing, \S\ref{S:methods} describes the methods used to construct the bitmaps and use them to identify files for queries, \S\ref{S:tests} presents several tests demonstrating the speed-up provided for querying the data, and \S\ref{S:discuss} summarizes the advantages and describes areas for application.

@@

%---------------------------------------------------------------------------------
% DOMAIN DECOMP
\subsection{Domain Partitioning Between Files}\label{SS:decomp}
A common analysis task is the selection of data within a subset of the full domain. If the data are split across multiple files either due to size constraints or parallel I/O, such selections require every file to be loaded and parsed. This can be very costly in terms of both the memory required to store the data and the time required to read it. However, if the contents of the files are mapped in advanced, only the files touched by the selection will need to be loaded. This is particularly effective for partitioning schemes that are localized within the domain. Figure \ref{fig:files} shows four examples of possible partitions of a two-dimensional domain equally across 8 files. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/files.png}
\caption{Examples of four different schemes for partitioning a 2D domain between 8 files. Each color represents a different file.}
\label{fig:files}
\end{center}
\end{figure}
%

Panel (a) is an example where random parts of the domain are contained within each file. In such a case, many files will need to be loaded for contiguous selections within the domain. In panel (b), the domain was split between the files along the $x$ dimension. Fewer files will need to be loaded for queries along the $y$-dimension, but contiguous selection in $x$ will still require a greater number of files. Panels (c) and (d) are both examples of partitioning the domain between the files along a space filling curve \citep[Morton and Hilbert curves respectively,][]{Morton1996,Hilbert1970}. These partitions have the greatest chance of limiting the number of files that must be loaded for a contiguous selection with slightly improved localization for the Hilbert curve.

Figure \ref{fig:selectors} shows examples of three selections within the above domain partitions. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector1.png}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector5.png}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector4.png}
\caption{Examples of file selection for four different domain partitions and three different selectors. The number of files above each images is the number of files that must be loaded in order to get all of the data within the selector.}
\label{fig:selectors}
\end{center}
\end{figure}
%
For the smallest selector (first row), the random domain decomposition (a) already requires half of the files to be loaded while more localized schemes require much fewer. Similarly, while the sliced domain partition (b), requires the fewest files to be loaded when the selector is oriented in the same direction as the slicing (second row), it requires all of the files when the selector is perpendicular to the slicing (third row). While some datasets may have information on the domain range covered by each file, the partitioning scheme used at output is often decided at runtime, can be system dependent, and may be imperfect.

Files are often partitioned for parallel I/O such that each processor outputs data on the portion of the domain it is responsible for. To limit the cost of communication between processors, the domain will be split across processors such that neighboring processors are responsible for neighboring parts of the domain. This means that, although the overall partitioning scheme may be known for a given dataset, the exact order of the files will be dependent on the configuration of the processors at runtime.

The partitioning can also be imperfect if the partition is not perfectly maintained. For instance, in astrophysical N-body simulations, it is possible for particles to travel from one processor's domain to another. In this case, the partition will only be perfect directly following an update to the domain decomposition.

In cases where the exact file organization is not known or imperfect, it is advantageous to map the files post-process in order to speed up selections for analysis. Although the same result can be achieved by re-sorting the data itself, creating the map is computationally less expensive than re-sorting the data and does not required write access, which may not be feasible as in the case of datasets shared by large, distributed communities.

%---------------------------------------------------------------------------------
% MORTON INDICES
\subsection{Morton Indices}
Morton ordering maps multidimensional data onto a one-dimensional space filling curve \citep{Morton1996}. This is done by breaking up the domain into cells where each cell's position within the $N$-dimensional domain can be described by $N$ integers. The Morton index of the cell is then created by interleaving the bits of the $N$ integers to create a single integer that fully describes the cell's position (see panel (b) Figure \ref{fig:zorder}). As seen in panel (a) of Figure \ref{fig:zorder}, ordering of the cells by their Morton indices forms a space filling Z-curve.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/zorder.png}
\caption{Example of 3rd order Morton curve. The bits of the $x$ and $y$ indices are interleaved to generate a single integer fully descriptive of it's location within the two-dimensional domain to within $1/2^{3}$th in each dimension.}
\label{fig:zorder}
\end{center}
\end{figure}
%

The precision of a single Morton index is only limited by the size of the integer used to store it. For instance, 64-bit Morton indices in 3 dimensions can be localized to $1/2^{21}$th of the domain in each dimension ($3\times21$\,bits = 63\,bits). If the domain is binarily divided into subcells to some order $k$ in each dimension (i.e. $2^{Nk}$ cells), coarser Morton indices can be obtained by simply masking lower bits. Morton ordering has been used to speed up quadtree construction \citep{Hjaltason2002}, nearest neighbor searches \citep{Connor2010}, and range queries \citep{Orenstein1984}. By recording the indices of the cells containing data from each file within a dataset, Morton indices can also be used to construct one-dimensional maps of an $N$-dimensional dataset.

%---------------------------------------------------------------------------------
% BITMAPS & EWAH
\subsection{Bitmaps \& EWAH Compression}
Bitmap indexes use the values of single bits to describe dataset properties. They are popular for use with large data warehouses \citep{Wu1998,Chan1998,Chan1999}, but have also been applied to scientific data from a diverse array of fields including geosciences \citep{Malensek2014}, earth sciences, rocket science \citep{Sinha2006,Sinha2007}, high-energy physics \citep{Stockinger2000}, and combustion \citep{Wu2003}.

In cases where data attributes can take on a finite set of values, one bitmap is constructed for each possible attribute value. Within the bitmap each bit specifies whether or not the corresponding data point has that value. In this way, queries for data with a single attribute value require considering only one bitmap and queries of multiple attributes/values can be done using boolean AND operations on the corresponding bitmaps. In the case of scientific data, which often contains floating point value attributes, such attributes must be binned prior to constructing the bitmaps \citep{Yu1998,Shoshani1999}. Here, Morton indices are used to bin N-dimensional floating point data along one-dimension and one bitmap is constructed for each file.

For each file within a dataset, the Morton indices touched by the data can then be stored in a bitmap index for future searches where the value of bit $j$ indicates whether or not Morton index $j$ is touched by the file in question. For Morton indexing of order $k$, this would result in a bitmap of length $2^{Nk}$ bits per file. This can become costly in terms of memory and the time required to perform bitmap operations. However, Enhanced Word-Aligned Hybrid (EWAH) compression can be used to limit these costs, particularly when the domain is densely or sparsely populated in localized regions \citep{Wu2001,Lemire2010,Kaser2016}.

An EWAH compressed bitmap will be smaller when there are long runs of identical values, with the smallest case being one in which either all or none of the bits are set. An uncompressed bitmap would require the same, maximum, amount of memory in both of these cases. The locality of Morton indices also takes advantage of the EWAH compression. If there are regions of the domain that are densely/sparsely populated, there Z-order space filling curve ensures that the bits denoting those regions will be adjacent, increasing the likelihood that there will be runs of identical (set/unset) bits and limiting the size of the compressed bitmaps.

%---------------------------------------------------------------------------------
% COLLISIONS
\subsection{Collisions}
It is possible that two files will contain data within the same Morton cell. This would mean that any time the data within that cell is accessed, both files would need to be loaded. Figure \ref{fig:collision} provides an example of collisions between two files. In panel (a) of Figure \ref{fig:collision}, purple cells are those that contain data from both files, a collision, for a 3rd order Morton index. Any selector that contained one of those cells would need to load all of the data from both files, even if it only selected part of the cell.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/collisions.png}
\caption{Examples of a collision between two files. The red points and blue points are contained by two different files. The larger grid in both panels denotes the boundaries of 3rd order Morton cells.  The cells containing points from either file are shaded accordingly such that cells containing points from both files are purple. The smaller grids within these cells on the right are the boundaries of 2nd order Morton cells refining the collisions.}
\label{fig:collision}
\end{center}
\end{figure}
%

@@

%---------------------------------------------------------------------------------
% GHOST ZONES
\subsection{Ghost Zones}
It is often the case that, in selecting a region, additional padding around the region should be included in the selection. This is particularly useful for algorithms that need information about neighboring points in the domain \addref. For Morton indices, this is straightforward as the indices neighboring Morton cells can be found by incrementing the bits corresponding to each dimension. We have included the ability to pad selectors with some number of Morton cells referred to as `ghost zones'. Those files that touch ghost zones, but not the selector itself are referred to below as `ghost files'.

@@

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% METHODS
\section{Methods}\label{S:methods}
The basic procedure for constructing the bitmap index is as follows:
\begin{enumerate}
\item {\bf Compute coarse indices.} For each file in the data set, read in the data and compute the indices of Morton cells at a given coarse order that are touched by data contained by that file. Store the coarse Morton indices in an EWAH compressed bitmap.
\item {\bf Find collisions.} Locate indices of coarse cells that are touched by data in more than one file (collisions) using bitwise operations. Record the collision indices in an EWAH compressed bitmap.
\item {\bf Compute refined indices.} For each file in the data set, read in the data and compute the indices Morton cells at a given refined order within coarse cells with collisions that are touched by data within that file. Store the refined Morton indices in a map from coarse Morton index to an EWAH compressed bitmap of refined Morton indices within that cell.
\item {\bf Output bitmaps.} Save EWAH compressed bitmaps for the coarse and refined indices contained with in each file in addition to the EWAH compressed bitmap containing the coarse indices with collisions.
\end{enumerate}

For large datasets and/or high order bitmaps, this can be a time consuming process. However, it must only be done once. For future selections, the bitmap can be quickly loaded and used to identify files in less time than would be required to load and query each file within the dataset individually. Selection using a loaded bitmap goes as follows:
\begin{enumerate}
\item {\bf Construct selector bitmap.} In the same way each file was mapped, the indices of Morton cells touched by the selector can be stored in a bitmap. This is done by checking for intersection of the selector with Morton cells at the order of the coarse bitmaps. For contiguous selectors, this can be done at lower order (parent) cells first and continued recursively until the order of the coarse bitmap is reached. 
\begin{itemize}
\item If a cell is completely within the selector, all of its child cells at the coarse order are added to the bitmap. 
\item If a cell intersects the edge of the selector, child cells at increasing orders are checked until the order of the coarse bitmaps are reached. 
\end{itemize}
If there is a collision between files within a coarse cell, a second order bitmap is constructed for the selector in the same manner.
\item {\bf Find files intersecting the selector.} Bitwise operations with the coarse file bitmaps can be used to efficiently identify files that intersect the selector within coarse cells. If the coarse cells within the intersection with a file all have collisions with other files, bitwise operations with the refined file bitmaps are then used to determine if the file is actually selected.
\end{enumerate}

@@

These procedures were implemented as part of the {\tt yt} python package \citep{Turk20d11a} in order to facilitate the analysis of large astrophysical N-body simulations. The open source EWAHBoolArray C++ package is used for implementing EWAH bitmaps \citep{Lemire2010,Kaser2016}.
\todo{more/separate code section...?}

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% TESTS
\section{Tests}\label{S:tests}
The utility of using Morton index bitmaps for mapping files to decrease query times was tested on artificial datasets containing $1024^3$ points in three dimensions, distributed between 512 files. For each test a Morton index bitmap was constructed for the dataset and used to identify files touched by cube shaped three-dimensional selectors. The performance was then assessed in terms of the number of files touched and the time required to identify the files touched. If fewer files are touched, fewer files will need to be loaded during analysis of a selected region and the overall fraction of time spent on I/O will be lower. If less time is required to identify the files touched by a given selector, more selections can be made using the same computational resources. This was done for varying index orders (\S\ref{SS:test_order}), selector sizes (\S\ref{SS:test_size}), and partitions of the domain across the files (\S\ref{SS:test_decomp}).

% INDEX ORDER
% TOTAL REFINEMENT
\subsection{Index Order}\label{SS:test_order}
\subsubsection{Overall Refinement}\label{SSS:test_order1}
The order of the Morton indices used to map the files determines the time required to identify files and the number of collisions that will occur between files. Higher order indices will result in fewer collisions, but will take longer to query, as seen in Figure \ref{fig:test_order1}. Six selectors of varying sizes and positions within the domain where used to identify files based on Morton index bitmaps of varying order. The test dataset was split across the files using a Hilbert curve of order 6 with 10\% scatter between Hilbert cells to simulate an imperfect domain decomposition as can occur if particle positions are updated and output prior to communication.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/vary_order1_np1024_nf512_or0.png}
\caption{Dependence of query time (top) and fraction of files selected/cells with collisions (bottom) on total refinement of Morton indices. The solid black lines correspond to the query times and files identified by just the selectors. The dashed black lines correspond to the query times and additional files selected when a ghost zone with the width of one Morton cell is added around the selectors. The solid blue line in the bottom panel shows the fraction of cells with collisions between files.}
\label{fig:test_order1}
\end{center}
\end{figure}
%

Below an order of 4, there are collisions between multiple files for every index, resulting in a larger number of files being identified. However, as the order increases, the number of collisions drops and the file count plateaus at $\sim25$\%. For indices of order 7, selection requires $>100\times$ the time that the same selection took for an order of 6, but there is no change in the number of files indicated. For this dataset, an order of 6 is sufficient to identify the minimal set of files touched by the selectors. This is the order of the domain decomposition for the file partitioning.

% SECONDARY REFINEMENT
\subsubsection{Collision Refinement}\label{SSS:test_order2}
Increasing the refinement of the primary index does so for the entire domain and can become costly in terms of the memory required to store the bitmap and the time required to perform operations. However, it is also possible to increase refinement by nesting a second Morton index within those cells of the primary index that contain collisions. The resulting increase in memory and operations cost for adding the second index is then proportional to the number of collisions within the primary bitmap. This means that using a second index is only advantageous over increasing the order of the first index in some cases. Figure \ref{fig:test_order2} shows the results for adding a secondary index of varying order with the overall refinement order of the bitmap (primary index order + secondary index order) held constant at 6. The test dataset and selectors applied were the same as in \S\ref{SSS:test_order1}.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/vary_order2_np1024_nf512.png}
\caption{Dependence of query time (top) and fraction of files selected/cells with collisions (bottom) on primary vs. secondary refinement. The solid black lines correspond to the query times and files identified. The dashed blue line in the bottom panel is the fraction of cells at the first index level that have collisions. The dotted blue line is the fraction of cells at the second index level that have collisions.}
\label{fig:test_order2}
\end{center}
\end{figure}
%

When the order of the second refined index is low, the first index is larger resulting in fewer cells with collisions at the first index and more at the second. The reverse is true when the order of second index is higher. As the overall order is held constant, the same number of files are identified regardless of the orders of the first and second indices. The time required to identify the files is minimized when cells within the first index become saturated with collisions. For higher order secondary indices, operations on the secondary bitmaps begin to be costly in terms of time for operations and memory consumption, but are still less expensive than in the case where only a single index is used. The optimal value for the orders of the first and second indexes will depend upon the dataset in question. The test dataset used here is uniformly distributed throughout the domain and does not need a high level of refinement at collisions. However, if a data set were less uniform with concentrations of points, the optimal order of the second index for performance may be higher.

>>>>

%---------------------------------------------------------------------------------
% ABSTRACT
\begin{abstract}
Current scientific datasets can reach unwieldy sizes that impede analysis and scientific discovery. We present a method for using Morton order bitmap indexes to map the parts of the full domain touched by files within a dataset containing discrete points. This technique takes advantage of domain partitioning across multiple files and limits the volume of data that must be loaded into memory for a single query. This method is tested on mock datasets containing $1024^3$ three-dimensional points split across 512 files according to different partitioning schemes. Using Morton order bitmap indexes for localized partitioning schemes like the Hilbert space-filling curve, the number of files that must be loaded to access data for a sub-section of the domain will scale with the size of the sub-section and can be identified in under 0.01\,s even when the entire domain is included. Mapping data files in this manner will dramatically decrease the time and memory required for loading data during analysis without needing to re-order the data on disk.
\end{abstract}

@@

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% INTRO
\section{Introduction}\label{S:intro}
Continuing advances in both hardware and software facilitate astrophysical datasets of growing complexity and size. The datasets produced by numerical simulations can currently reach sizes of a $\sim$100\TB{ }split across hundreds of files \citep[e.g.][]{Croft2015}. For even simple analysis tasks, the cost of incrementally reading datasets this large into memory is quite high. This problem is not limited to theoretical work. During operations the Large Synoptic Survey Telescope (LSST) will produce 15\TB{ }of data each night \citep{Juric2015a}. In order to analyze such large datasets, we need innovative techniques for quickly indexing and selecting data without loading the entire dataset into memory. We present a technique for using Morton bitmap indexes to map files and accelerate data analysis. \S\ref{S:theory} provides background information on the file indexing scheme presented here, \S\ref{S:methods} describes the methods used to construct the indexes and use them to identify files for queries, \S\ref{S:tests} presents several tests demonstrating the speed-up provided for querying the data, and \S\ref{S:discuss} summarizes the advantages, describes areas for application, and proposes areas for future improvement.

@@

%---------------------------------------------------------------------------------
% DOMAIN DECOMP
\subsection{Domain Partitioning Between Files}\label{SS:decomp}
A common analysis task is the selection of data within a subset of the full domain. If the dataset is split across multiple files, either due to size constraints or to allow for parallel I/O, such selections require every file to be loaded and parsed in order to assemble all of the data within the selection criteria. This process can be very costly in terms of both the memory required to store the data and the time required to read each file. However, if the contents of the files are mapped in advanced, only the files touched by the selection will need to be loaded. This is particularly effective for partitioning schemes that are localized within the domain. If each file contains data that are localized to one part of the domain, selections of contiguous sub-sections within the domain will require fewer files to be loaded. Figure \ref{fig:files} shows four examples of possible partitions of a two-dimensional spatial domain split equally between 8 files. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/files.png}
\caption{Examples of four different schemes for partitioning a 2D domain between 8 files. Each color represents a different file.}
\label{fig:files}
\end{center}
\end{figure}
%

Panel (a) is an example where random parts of the domain are contained within each file. In such a case, many files will need to be loaded for contiguous selections within the domain. In panel (b), the domain was split between the files along the $x$ dimension. Fewer files will need to be loaded for queries along the $y$-dimension, but contiguous selection in $x$ will still require a greater number of files since the partition is not well localized in that dimension. Panels (c) and (d) are both examples of partitioning the domain between the files along a space filling curve \citep[Morton and Hilbert curves respectively;][]{Morton1996,Hilbert1970}. These partitions have the greatest chance of limiting the number of files that must be loaded for a contiguous selection with slightly improved localization for the Hilbert curve.

Figure \ref{fig:selectors} shows examples of three selections within the above domain partitions. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector1.png}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector5.png}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/selector4.png}
\caption{Examples of file selection for four different domain partitions and three different shaded selectors. The number of files above each images is the number of files that must be loaded in order to get all of the data within the selected region.}
\label{fig:selectors}
\end{center}
\end{figure}
%
For the smallest selector (first row), the random domain decomposition (a) already requires half of the files to be loaded while more localized schemes require much fewer. Similarly, while the sliced domain partition (b), requires the fewest files to be loaded when the selector is oriented in the same direction as the slicing (second row), it requires all of the files when the selector is perpendicular to the slicing (third row). While some datasets may have information on the domain range covered by each file, the partitioning scheme used for simulation output is often decided at runtime, can be system dependent, and may be imperfect.

Files are often partitioned for parallel I/O such that each processor outputs data on the portion of the domain it is responsible for processing. To limit the cost of communication between processors, the domain will be split across processors such that neighboring processors are responsible for neighboring parts of the domain. This means that, although the overall partitioning scheme may be known for a given dataset, the exact order of the files will be dependent on the configuration of the processors at runtime.

The partitioning can also be imperfect if the domain decomposition is not perfect at the time of output. For instance, in astrophysical N-body simulations, it is possible for particles to travel from one processor's domain to another. In this case, the partition will only be perfect directly following an update to the domain decomposition.

In cases where the exact file organization is not known or imperfect, it is advantageous to map the files post-process in order to speed up selections for analysis. Although the same result can be achieved by re-sorting the data itself, creating the map can be less computationally less expensive than re-sorting the data, can be saved for use with multiple selections, and does not required write access, which may not be feasible as in the case of datasets shared by large, distributed communities.

%---------------------------------------------------------------------------------
% MORTON INDICES
\subsection{Morton Indices}
Morton ordering maps multidimensional data onto a one-dimensional space filling curve \citep{Morton1996}. This is done by breaking up the domain into cells where each cell's position within the $N$-dimensional domain can be described by $N$ integers. The Morton index of the cell is then created by interleaving the bits of the $N$ integers to create a single integer that fully describes the cell's position (see panel (b) Figure \ref{fig:zorder}). As seen in panel (a) of Figure \ref{fig:zorder}, ordering of the cells by their Morton indices forms a space filling Z-curve.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/zorder.png}
\caption{Example of 3rd order Morton curve in two dimensions. The bits of the $x$ and $y$ indices are interleaved to generate a single integer that fully describes the cell's location within the two-dimensional domain to within $1/2^{3}$th of the domain in each dimension.}
\label{fig:zorder}
\end{center}
\end{figure}
%

The precision of a single Morton index is only limited by the size of the integer used to store it. For instance, 64-bit Morton indices in 3 dimensions can be localized to $1/2^{21}$th of the domain in each dimension ($3\times21$\,bits = 63\,bits). If the domain is binarily divided into subcells to some order $k$ in each dimension (i.e. $2^{Nk}$ cells), coarser Morton indices can be obtained by simply masking lower bits. Morton ordering has been used to speed up quadtree construction \citep{Hjaltason2002}, nearest neighbor searches \citep{Connor2010}, and range queries \citep{Orenstein1984}. By recording the indices of the cells containing data from each file within a dataset, Morton indices can also be used to construct one-dimensional maps of an $N$-dimensional dataset that can be represented as bitmaps.

%---------------------------------------------------------------------------------
% BITMAPS & EWAH
\subsection{Bitmaps \& EWAH Compression}
Bitmap indexes use the values of single bits within an array of bits to describe dataset properties. This form requires minimal memory and can be filtered using computationally inexpensive boolean operations. Bitmap indexes have long been popular for use with large data warehouses \citep{Wu1998,Chan1998,Chan1999}. However, as scientific datasets have become larger and more complex, they have also begun to gain traction in a diverse array of scientific fields including geosciences \citep{Malensek2014}, earth sciences, rocket science \citep{Sinha2006,Sinha2007}, high-energy physics \citep{Stockinger2000}, and combustion \citep{Wu2003}.

In cases where data attributes can take on a finite set of values, one bitmap is constructed for each possible attribute value. Within the bitmap each bit specifies whether or not the corresponding data point has that value. In this way, queries for data with a single attribute value require consulting only one bitmap and queries of multiple attributes/values can be done using boolean AND operations on the corresponding bitmaps. In the case of scientific data, which often contains floating point value attributes, the attributes must be binned prior to constructing the bitmaps \citep{Yu1998,Shoshani1999,Stockinger2004}. Here, Morton indices are used to bin N-dimensional floating point data onto one-dimension. As a result, each file can be described by one bitmap.

For each file within a dataset, the Morton indices touched by the data within that file can then be stored in a bitmap index for future searches where the value of bit $j$ indicates whether or not Morton index $j$ is touched by the file in question. For Morton indexing of order $k$, this would result in a bitmap of length $2^{Nk}$ bits per file. For large bitmaps, this can become costly in terms of memory and the time required to perform bitmap operations. However, Enhanced Word-Aligned Hybrid (EWAH) compression can be used to limit these costs, particularly when the domain is densely or sparsely populated in localized regions \citep{Wu2001,Lemire2010,Kaser2016}.

An EWAH compressed bitmap will be smaller when there are long runs of identical values. This means that an EWAH compressed bitmap will be smallest if either all or none of its bits are set. An uncompressed bitmap would require the same, maximum, amount of memory in both of these cases. The locality of Morton indices takes advantage of the EWAH compression. If there are regions of the domain that are densely/sparsely populated, there Z-order space filling curve ensures that the bits denoting those regions will be adjacent, increasing the likelihood that there will be runs of identical (set/unset) bits and limiting the size of the compressed bitmaps.

%---------------------------------------------------------------------------------
% COLLISIONS
\subsection{Collisions}
It is possible that two files will contain data within the same Morton cell. This would mean that any time that cell is touched by a selection, both files would need to be loaded even if the selection only touches data from one of the files. Figure \ref{fig:collision} provides an example of collisions between two files. In panel (a) of Figure \ref{fig:collision}, purple cells are those that contain data from both files, a collision, for a 3rd order Morton index. Any selector that contained one of those cells would need to load all of the data from both files, even if it only selected part of the cell.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/collisions.png}
\caption{Examples of a collision between two files. The red points and blue points are contained by two different files. The larger grid in both panels denotes the boundaries of 3rd order Morton cells.  The cells containing points from either file are shaded accordingly such that cells containing points from both files are purple. The smaller grids within these cells on the right are the boundaries of 2nd order Morton cells refining the collisions.}
\label{fig:collision}
\end{center}
\end{figure}
%

@@

%---------------------------------------------------------------------------------
% GHOST ZONES
\subsection{Ghost Zones}
It is often the case that, in selecting a region, additional padding around the region should be included in the selection. This is particularly useful for algorithms that need information about neighboring points in the domain \citep[e.g. gas properties in simulations using Smoothed Particle Hydrodynamics; SPH;][]{Springel2001,Springel2005b}. For Morton indices, this is straightforward as the indices neighboring Morton cells can be found by incrementing the bits corresponding to each dimension. We have included the ability to pad selectors with some number of Morton cells referred to as `ghost zones'. Those files that touch ghost zones, but not the selector itself are referred to below as `ghost files'.

@@

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% METHODS
\section{Methods}\label{S:methods}
The basic procedure for constructing the bitmap index is as follows:
\begin{enumerate}
\item {\bf Compute coarse indices.} For each file in the data set, read in the data and compute the indices of Morton cells at a given coarse order that are touched by data contained within that file. These coarse indices are then stored by setting the corresponding bits in an EWAH compressed bitmap.
\item {\bf Find collisions.} The indices of coarse cells that are touched by data in more than one file (collisions) are located using bitwise operations on the file bitmaps. These indices are also stored in an EWAH compressed bitmap.
\item {\bf Compute refined indices.} For each file in the data set, read in the data and compute the indices of Morton cells at a given refined order within coarse cells with collisions that are touched by that file. These refined indices are stored in a map from coarse Morton index to an EWAH compressed bitmap of refined Morton indices within that cell.
\item {\bf Output bitmaps.} The EWAH compressed bitmaps for the coarse indices, refined indices, and collisions are saved to a single file.
\end{enumerate}

For large datasets and/or high levels of refinement, this can be a time consuming process. However, it must only be done once. For future selections, the bitmap can be quickly loaded and used to identify files in less time than would be required to load and query each file within the dataset individually. Selection using a loaded bitmap goes as follows:
\begin{enumerate}
\item {\bf Construct selector bitmap.} In the same way each file was mapped, the indices of Morton cells touched by the selector are stored in a bitmap. This is done by checking for intersection of the selector with Morton cells at the order of the coarse bitmaps. For contiguous selectors, this is done at lower order (parent) cells first and continued recursively until the order of the coarse bitmap is reached. 
\begin{itemize}
\item If a cell is completely within the selector, all of its child cells at the coarse order are added to the bitmap. 
\item If a cell intersects the edge of the selector, child cells at increasing orders are checked until the order of the coarse bitmaps are reached. If the cell is at the coarse order and there is a collision between two files, a refined bitmap is the constructed for the selector in the same manner.
\end{itemize}
\item {\bf Find files intersecting the selector.} Bitwise operations with the coarse file bitmaps are then used to efficiently identify files that intersect the selector within coarse cells. If the coarse cells within the intersection with a file all have collisions with other files, bitwise operations with the refined file bitmaps are then used to determine if the file is selected at the order of the refined index.
\end{enumerate}

@@

%---------------------------------------------------------------------------------
%---------------------------------------------------------------------------------
% TESTS
\section{Tests}\label{S:tests}
The utility of using Morton index bitmaps for mapping files to decrease query times was tested on artificial N-body simulation datasets containing $1024^3$ points in three dimensions, distributed between 512 files. For each test a Morton index bitmap was constructed for the dataset and used to identify files touched by cube shaped three-dimensional selectors. The performance is assessed in terms of the number of files identified and the average time required to identify them across 10 runs. If fewer files are touched, fewer files will need to be loaded during analysis of a selected region and the overall fraction of time spent on I/O will be lower. If less time is required to identify the files touched by a given selector, more selections can be made using the same computational resources. This was done for varying index orders (\S\ref{SS:test_order}), selector sizes (\S\ref{SS:test_size}), and partitions of the domain between files (\S\ref{SS:test_decomp}).

% INDEX ORDER
% TOTAL REFINEMENT
\subsection{Index Order}\label{SS:test_order}
\subsubsection{Overall Refinement}\label{SSS:test_order1}
The order of the Morton indices used to map the files determines the time required to identify files and the number of collisions that will occur between files. Higher order indices will result in fewer collisions, but will take longer to query, as seen in Figure \ref{fig:test_order1}. Six selectors of varying sizes and positions within the domain where used to identify files based on Morton index bitmaps of varying order. The test dataset was split across the files using a Hilbert curve of order 6 with 10\% scatter between Hilbert cells to simulate an imperfect domain decomposition as can occur if particle positions are updated and output prior to updating the domain decomposition.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/vary_order1_np1024_nf512_or0.png}
\caption{Dependence of query time (top) and fraction of files selected/cells with collisions (bottom) on the total refinement of the bitmap index. The solid black lines correspond to the query times and files identified by just the selectors. The dashed black lines correspond to the query times and additional files selected when a ghost zone with the width of one Morton cell is added around the selectors. The solid blue line in the bottom panel shows the fraction of cells with collisions between files.}
\label{fig:test_order1}
\end{center}
\end{figure}
%

Below a bitmap index order of 4, there are collisions between multiple files within every cell, resulting in a larger number of files being identified. However, as the order increases, the number of collisions drops and the file count plateaus at $\sim25$\%. For a 7th order bitmap index, selection requires $>100\times$ the time that the same selection took using a 6th order index, but there is no change in the number of files indicated. A 6th order index is sufficient to identify the minimal set of files touched by the selectors in this case because the dataset was partitioned between the files along a 6th order Hilbert space filling curve. While it is generally true that the time required to identify files using a bitmap index will increase exponentially with the order of the index, the order of the index that results in the minimal number of files for any dataset will depend upon how the domain is partitioned between files (see \S\ref{SS:test_decomp}).

% SECONDARY REFINEMENT
\subsubsection{Collision Refinement}\label{SSS:test_order2}
Increasing the refinement of the primary index does so for the entire domain and, as seen in \S\ref{SSS:test_order1}, can become costly in terms of the memory required to store the bitmap and the time required to perform operations. However, it is also possible to increase refinement by nesting a second Morton bitmap index within those cells of the primary index that contain collisions. The increase in memory and operations cost that results from adding a second index is then proportional to the number of collisions within the primary bitmap. Figure \ref{fig:test_order2} shows the results for adding a secondary index of varying order with the overall refinement order of the index (primary index order + secondary index order) held constant at 6. The test dataset and selectors applied were the same as in \S\ref{SSS:test_order1}.
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\columnwidth,keepaspectratio]{../images/vary_order2_np1024_nf512.png}
\caption{Dependence of query time (top) and fraction of files selected/cells with collisions (bottom) on primary vs. secondary refinement. The solid black lines correspond to the query times and files identified. The dashed blue line in the bottom panel is the fraction of cells at the first index level that have collisions. The dotted blue line is the fraction of cells at the second index level that have collisions.}
\label{fig:test_order2}
\end{center}
\end{figure}
%

When the order of the second refined index is low, the first index is larger resulting in fewer cells with collisions at the first index and more at the second. The reverse is true when the order of second index is higher. As the overall order is held constant, the same number of files are identified regardless of the orders of the first and second indexes. The time required to identify the files is minimized when cells within the first index become saturated with collisions. For higher order secondary indices, the EWAH compressed bitmaps used to refine collisions at the coarse index order are larger, resulting in an increase in the time required for query operations and memory consumption. However, selections using higher order secondary indices still require less time than in the case where only a single index is used.

The optimal value for the orders of the first and second indexes will depend upon the dataset in question. The density of data points within the test dataset used here is relatively uniform throughout the domain and does not need a high level of refinement at collisions. However, if a data set were less uniform with concentrations of points, the optimal order of the second index for performance may be higher.
